% !TEX root = ../main.tex

\begin{achievements}

\subsection*{学术论文}

\begin{bibliolist}{00}
  \item \textbf{Xiupu Lang}, Lin Sun, Peiqi Huang, Jinming Liu, Boming Zhu, Xiaojian Gao, Lin Gui, and Haopeng Chen. 2025. Incentivizing Cooperation for Handover Strategies in LEO Constellations via Fairness-guided MARL. ACM Trans. Internet Technol. Just Accepted (August 2025). 
  \item \textbf{Lang, X.}, Gui, L. (2021). Design of Southbound Interfaces in Heterogeneous Software-Defined Satellite Networks. In: Yu, Q. (eds) Space Information Network. SINC 2020. Communications in Computer and Information Science, vol 1353. Springer, Singapore. 
  \item \textbf{Lang, X.}, Zhang, Q., Gui, L., Hao, X., Chen, H. (2020). A Novel Topology Design Method for Multi-layered Optical Satellite Networks. In: Yu, Q. (eds) Space Information Networks. SINC 2019. Communications in Computer and Information Science, vol 1169. Springer, Singapore.
  \item Hao Li, \textbf{Xiupu Lang}, Lin Gui, "Research on intelligent network control mechanism and protocol verification of spatial information network," Proc. SPIE 11617, International Conference on Optoelectronic and Microelectronic Technology and Application, 116172Y (4 December 2020)
  \item A. Wei, H. Yu, \textbf{X. Lang} and B. Yang, "Dynamic Controller Placement for Software-Defined LEO Network Using Deep Reinforcement Learning," 2021 7th International Conference on Computer and Communications (ICCC), Chengdu, China, 2021, pp. 1314-1320.
  \item Q. Zhang, L. Gui, S. Zhu and \textbf{X. Lang}, "Task Offloading and Resource Scheduling in Hybrid Edge-Cloud Networks," in IEEE Access, vol. 9, pp. 85350-85366, 2021.
  \item S. Zhu, L. Gui, D. Zhao, N. Cheng, Q. Zhang and \textbf{X. Lang}, "Learning-Based Computation Offloading Approaches in UAVs-Assisted Edge Computing," in IEEE Transactions on Vehicular Technology, vol. 70, no. 1, pp. 928-944, Jan. 2021.
  \item S. Zhu, L. Gui, Q. Zhang and \textbf{X. Lang}, "Joint Client Association and UAV Scheduling in Cache-Enabled UAV-Assisted Vehicular Networks," 2021 IEEE 93rd Vehicular Technology Conference (VTC2021-Spring), Helsinki, Finland, 2021, pp. 1-6.
  \item Shichao Zhu, Lin Gui, Nan Cheng, Qi Zhang, Fei Sun, and \textbf{Xiupu Lang}. 2020. UAV‐enabled computation migration for complex missions: A reinforcement learning approach. IET Communications 14, 15 (September 2020), 2472–2480.
\end{bibliolist}

\subsection*{专利}
\begin{bibliolist}{00}
  \item 郎修璞，“一种保障高优先级时敏业务可靠传输的自组织网络”，专利申请号ZL202111027479.4
  \item 郎修璞，“基于SDN的卫星网络管控协议半实物试验装置”，专利申请号ZL202011608580.4
  \item 郎修璞，“基于分布式TDMA的MAC传输方法以及无线自组织网络系统”，专利申请号ZL201910957180.5
  
\end{bibliolist}

\end{achievements}
